---
title: "Working with Color Palettes"
format: html
vignette: >
  %\VignetteIndexEntry{Color Palettes}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 8, fig.height = 5)
```

In the [Getting started with ROvis vignette](getting-started-ROvis.html) you learned how to explore colors and how to use the `ro_color()` function in your ggplots. You also learned how to use the categorical color palette. However, ROvis also has functions to create color sequences to help you with visualizing ordinal and continuous data.

## Loading data

First, we need to load the required packages.
```{r libraries, message=FALSE}
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(ROvis.utils) # nolint: missing_package_linter
```

Second, we generate simulated data that mimics the structure of COVID-19 test data. We will use data for January 2022 for example graphs.
```{r data}
# Set seed for reproducibility
set.seed(123)

# Generate data for multiple regions used in the examples
dat_test_cov <- expand_grid(
  Date_of_statistics = seq(as_date("2022-01-01"), as_date("2022-01-31"), by = "day"),
  Security_region_name = c(
    "Brabant-Noord",
    "Brabant-Zuidoost",
    "Midden- en West-Brabant",
    "Noord- en Oost-Gelderland",
    "Gelderland-Midden",
    "Gelderland-Zuid",
    "Onbekend"
  )
) |>
  mutate(
    # Calculate day number (1-31) for upward trend
    day_num = as.numeric(Date_of_statistics - as_date("2022-01-01")) + 1,
    # Generate test numbers with slight variation
    Tested_with_result = as.integer(round(rnorm(n(), mean = 1800, sd = 50))),
    # Positive tests with steep upward trend: start at ~15% and increase to ~50%
    positive_rate = 0.15 + (day_num - 1) * (0.50 - 0.15) / 30,
    Tested_positive = as.integer(round(
      Tested_with_result * positive_rate * rnorm(n(), mean = 1, sd = 0.05)
    ))
  ) |>
  select(Date_of_statistics, Security_region_name, Tested_positive) |>
  mutate(Security_region_name = factor(Security_region_name)) |>
  as_tibble()
```

The first 5 rows look like this:
```{r data show}
dat_test_cov |>
  head(5)
```


## Coloring ordinal data

### Create a color sequence
Using a color sequence can be useful for coloring ordinal data. We can use the `ro_color_seq` function to create a color sequence. Let's create a sequence from light to dark 'hemelblauw':

```{r color seq hemelblauw}
ro_color_seq(ro_color("hemelblauw_tint15"), ro_color("hemelblauw"), 5)
```

We can visually inspect our sequence with `show_colors`:
```{r show color seq hemelblauw}
ro_show_colors(
  "custom",
  ro_color_seq(ro_color("hemelblauw_tint15"), ro_color("hemelblauw"), 5)
)
```

We can also create a color sequence from one color flowing into another color. For example start with 'robijnrood' and end with 'hemelblauw':
```{r color seq hemelblauw to robijn}
ro_color_seq(ro_color("hemelblauw"), ro_color("robijnrood"), 8)
```

We can visually inspect our sequence with `show_colors`:
```{r show color seq hemelblauw to robijn}
ro_show_colors(
  "custom",
  ro_color_seq(ro_color("hemelblauw"), ro_color("robijnrood"), 8)
)
```


### How to use `ro_color_seq()` in `ggplot` functions

To demonstrate the use of a color sequence in a ggplot graph, we select the security regions in "Brabant" using `filter`. We start by creating a plot without specifying colors. Thus, the resulting plot shows standard colors of ` ggplot`.

*(Please note: for the sake of this example, categorical data is used. However, color gradients are especially powerful in visualizations of ordinal or continuous data).*

```{r tab_brab_geld}
tab_brab_geld <- dat_test_cov |>
  filter(
    Security_region_name %in%
      c(
        "Midden- en West-Brabant",
        "Brabant-Noord",
        "Brabant-Zuidoost",
        "Noord- en Oost-Gelderland",
        "Gelderland-Midden",
        "Gelderland-Zuid",
        "Onbekend"
      )
  ) |>
  # For the sake of the example, other levels get dropped from `Security_region_name`.
  mutate(Security_region_name = droplevels(Security_region_name))
```


```{r fig_brab_geld_without_theme, message=FALSE}
fig_brab_geld <- tab_brab_geld |>
  ggplot(aes(
    x = Date_of_statistics,
    y = Tested_positive,
    color = Security_region_name
  )) +
  geom_line(linewidth = 1) +
  geom_point()

fig_brab_geld
```


We can apply the RIVM colors by using `ro_color_seq()`. We will also make the legend items appear in three rows for readibility:

```{r add-color-seq-to-fig}
fig_brab_geld +
  scale_color_manual(
    values = c(ro_color_seq(ro_color("hemelblauw"), ro_color("robijnrood"), 7)),
    guide = guide_legend(nrow = 3)
  )
```


### How to use `ro_color_seq_named()` in `ggplot` functions

When including an explicit "Not available" category (e.g., "NA"), it can be helpful to use the `ro_color_seq_named()` function. This function works similarly to `ro_color_seq()`, but it directly assigns colors to the specified category names. This allows for more precise control over color mapping in ordered visualizations.

```{r plot-brab-geld-with-named, message=FALSE}
tab_brab_geld |>
  ggplot(aes(
    x = Date_of_statistics,
    y = Tested_positive,
    color = Security_region_name
  )) +
  geom_line(linewidth = 1) +
  geom_point() +
  scale_color_manual(
    name = "Security region name",
    values = ro_color_seq_named(
      cat_names = levels(tab_brab_geld$Security_region_name),
      low_col = "lichtblauw",
      high_col = "hemelblauw",
      NA_cat_name = "Onbekend",
      NA_cat_col = "rood"
    ),
    guide = guide_legend(nrow = 3)
  )
```
